---
title: "Data Preparation"
format: html
editor: visual
---

## Setup

```{r}
library(readxl)
library(dplyr)
library(stringr)
library(writexl)
```

## Load the raw data

```{r}
raw <- read_excel("09_Sales_Revenue_Performance.xlsx", sheet = "1_RAW_DATA")

dim(raw)
```

## Exploratory profiling

```{r}
glimpse(raw)
```

### Missing values

```{r}
colSums(is.na(raw))
```

### Duplicates and messy categories

```{r}
cat("Exact-duplicate rows:", sum(duplicated(raw)), "\n\n")

cat("Region values:\n");         print(unique(raw$Region))
cat("\nPipeline stage values:\n"); print(unique(raw$Pipeline_Stage))
cat("\nNegative discounts:", sum(raw$`Discount_%` < 0, na.rm = TRUE), "\n")
```

## Cleaning

Each fix below addresses one issue found above. The steps are chained so the transformation reads as a single, reproducible pipeline.

```{r}
clean <- raw %>%

  # Remove exact-duplicate rows
  distinct() %>%

  # Standardise Region into canonical labels (accents kept to match the
  # dashboard), mapping the stray city to its region
  mutate(
    Region = str_trim(Region),
    Region = case_when(
      str_to_lower(Region) %in% c("miền bắc", "mien bac")               ~ "Miền Bắc",
      str_to_lower(Region) %in% c("miền nam", "mien nam", "ho chi minh") ~ "Miền Nam",
      str_to_lower(Region) %in% c("miền trung", "mien trung")           ~ "Miền Trung",
      str_to_lower(Region) %in% c("tây nguyên", "tay nguyen")           ~ "Tây Nguyên",
      TRUE ~ Region
    )
  ) %>%

  # Standardise Pipeline_Stage — collapse spelling variants
  mutate(
    Pipeline_Stage = str_trim(Pipeline_Stage),
    stage_key = str_to_lower(str_replace_all(Pipeline_Stage, " ", "")),
    Pipeline_Stage = case_when(
      stage_key == "closedwon"  ~ "Closed Won",
      stage_key == "closedlost" ~ "Closed Lost",
      TRUE ~ Pipeline_Stage
    )
  ) %>%
  select(-stage_key) %>%

  # Correct invalid negative discounts
  mutate(`Discount_%` = if_else(`Discount_%` < 0, abs(`Discount_%`), `Discount_%`)) %>%

  # Replace junk numeric values in Industry
  mutate(
    Industry = case_when(
      is.na(Industry)                   ~ NA_character_,
      str_detect(Industry, "^[0-9.]+$") ~ "Other",
      TRUE ~ Industry
    )
  )
```

### Handling missing values

```{r}
colSums(is.na(clean))
```

## Validation

Two checks confirm the financial logic is internally consistent: net revenue should never exceed contract value, and gross profit should equal net revenue minus cost of sale.

```{r}
chk_net <- sum(clean$Net_Revenue_VND > clean$Contract_Value_VND, na.rm = TRUE)
chk_gp  <- sum(abs((clean$Net_Revenue_VND - clean$Cost_of_Sale_VND) -
                     clean$Gross_Profit_VND) > 1000, na.rm = TRUE)

cat("Rows where Net > Contract (expect 0):", chk_net, "\n")
cat("Gross profit reconciliation errors (expect 0):", chk_gp, "\n")
cat("Final dimensions:", nrow(clean), "rows x", ncol(clean), "columns\n\n")
cat("Regions after cleaning:\n"); print(unique(clean$Region))
```

Both checks pass: the money columns are consistent, and `Region` now holds exactly four clean, canonical values.

## Export

```{r}
write_xlsx(clean, "sales_data_cleaned.xlsx")
write.csv(clean, "sales_data_cleaned.csv", row.names = FALSE, fileEncoding = "UTF-8", na = "")
```
